Citation Analysis of Computer Science Publications

### 2019-12-05 Scientometrics/comp.tex
update

\title{Citation Analysis of Computer Science Publications%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
\subtitle{Relationship To Other Fields}

@@

As far back as 1979, Salton and Bergmark conducted a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles) in which they made the point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas noted were (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Computing has grown and evolved significantly since 1974. Progressing beyond Salton and Bergmark's historical triad of theoretical foundations; hardware and computer systems; and software, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification does address relationships with other fields under Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from adjacent fields, to the CCS classification does not exist.

@@

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs = 200$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 22,232 clusters containing 84,591 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned as the maximum edge weight of all edges between the nodes in these two clusters. Edges were arranged in descending order. The first pair of clusters were merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (figure), and resulted in 665 clusters. The top 20 clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using Graclus, a spectral clustering technique that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. Considering the caveats associated with generating \emph{comp}, namely, the coverage limitations of the DBLP dataset (ii) cross-matching to Scopus (iii) journal based classification (iv) clustering, we constructed article level clusters of this computer science dataset at a sufficiently high level of aggregation to avoid cognitive challenge and cross-matched them to disciplinary classification by journal, the Scopus ASJC classification. We only considered Scopus ASJC minor subject area categories that accounted for at least 10\% of the publications in each cluster.

@@

From the other perspective, cluster 17 was the most diverse and contained publications annotated with 8 minor subject area labels, of which Biochemistry, Chemistry(all), Genetics, and Molecular Biology were the most distant from computer science while Statistics \& Probability, Computational Theory \& Mathematics, and Computational Mathematics are closer.  These data suggest that areas central to computer science in 2019 (Salton and Bergmark's triad) are more likely to be found in multiple clusters (hardware, software, and theory), and that biology, chemistry, and statistics are the most highly represented fields proximal to computer science.  These results also suggest that the actual structure of communication is at variance with journal-based classification, also noted by Waltman and van Eck, ``...an indication that traditional disciplines such as those just mentioned only partly reflect the actual organization of today's scientific research."~\cite{waltman_new_2012}. The aphorism of Porter, Cohen, Roessner, and Perreault (2007), `The discrepancies between disciplinary roots and research realities loom large'~\cite{porter_measuring_2007} also invokes itself. A second inference is that a correlation appears between some cases of journal-based classification and article clusters in terms of subject matter. Thirdly, that some areas of computer science are relatively ubiquitous, e.g. computer applications. Last, that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17 (ibid) and cluster 18 (artificial intelligence, cognitive neuroscience, and neurology).

@@

Co-citation, the frequency with which a pair of articles is cited by other articles, was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973} provides insight into the emergence of new ideas derived from the association of previously independent ones. Co-citation, thus provides some degree of evolutionary insight across disciplines and sub-disciplines and has been used for clustering publications~\cite{boyack_co-citation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}. To provide an alternate view of these DBLP data we also constructed clusters using co-citation as below and as described in Methods.

@@

 \begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.3]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{Summary of DBLP data cross-matched with Scopus. 2,685,356 publications from DBLP were cross-matched with Scopus and then grouped by the 27 major subject areas in the ASJC (Scopus) classification. The largest number of publications are contributed by Computer Science; Mathematics; and Engineering and then by Social Sciences; Physics and Astronomy; and Biochemistry, Genetics, and Molecular Biology. Publications were further annotated with respect to being either articles \emph{(ar)} or conference proceedings \emph{(cp)}. For this dataset, the major subject areas of Computer Science; Engineering; and Mathematics contributed the most publications while Pharmacology, Toxicology, and Pharmaceutics; Nursing, Veterinary and Dentistry contributed the least.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}

>>>>

\title{Citation Analysis of Computer Science Publications}
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}

\subtitle{Relationship To Other Fields}

@@

Noting that that the scientific literature serves a rich source of information to study the structure and historical development of a field, Salton and Bergmark conducted a historical study  of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles)~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas noted were (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Computing has grown and evolved significantly since 1974. Progressing beyond Salton and Bergmark's historical triad of theoretical foundations; hardware and computer systems; and software, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification also addresses relationships with other fields under the category of Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from adjacent fields, to the CCS classification does not exist.

@@

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 22,232 clusters containing 84,591 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned as the maximum edge weight of all edges between the nodes in these two clusters. Edges were arranged in descending order. The first pair of clusters were merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (figure).

@@

\emph{Clustering by Direct Citation.} Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using Graclus, a spectral clustering technique that has previously been applied to citation data~\cite{graclus_2007,subelj_clustering_2016}. Considering the caveats associated with generating \emph{comp}, namely, the coverage limitations of the DBLP dataset (ii) cross-matching to Scopus (iii) journal based classification (iv) clustering, we constructed article level clusters of this computer science dataset at a sufficiently high level of aggregation to avoid cognitive challenge and cross-matched them to disciplinary classification by journal, the Scopus ASJC classification. We only considered Scopus ASJC minor subject area categories that accounted for at least 10\% of the publications in each cluster.

@@

From the other perspective, cluster 17 was the most diverse and contained publications annotated with 8 minor subject area labels, of which Biochemistry, Chemistry(all), Genetics, and Molecular Biology were the most distant from computer science while Statistics \& Probability, Computational Theory \& Mathematics, and Computational Mathematics are closer.  These data suggest that areas central to computer science in 2019 (Salton and Bergmark's triad) are more likely to be found in multiple clusters (hardware, software, and theory), and that biology, chemistry, and statistics are the most proximal to computer science. [What about mathematics?]  A second inference is that, in a few cases journal-based classification and article clusters are tightly related in terms of article content (Hardware and Architecture). Thirdly, that some areas of computer science are relatively ubiquitous, e.g. computer applications. Last, that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17 (ibid) and cluster 18 (artificial intelligence, cognitive neuroscience, and neurology).

@@

\emph{Clustering by Co-citation.} Co-citation, the frequency with which a pair of articles is cited by other articles, was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973} provides insight into the emergence of new ideas derived from the association of previously independent ones. Co-citation, thus provides some degree of evolutionary insight across disciplines and sub-disciplines and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}. For an alternate view of these DBLP data we constructed clusters using co-citation by first applying variable level clustering to the 2,685,356 articles in \emph{comp} and then performing top-down agglomerative clustering on the resultant 10,298 (of 22,232 clusters) containing greater than two nodes per cluster. (Methods). We performed 600 iterations to the 10,298 clusters further limited to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}). The 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
 \includegraphics[scale=0.5]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{High Level Disciplinary Composition of Publications in Clusters Generate by Variable Level Clustering. 84,591 nodes from were matched to the 4 subject areas, 
27 major subject areas, and 334 minor subject areas in the Scopus ASJC classification. Since 82.5\% of these nodes were assigned more to than one minor subject category, 
we assigned fractional counts for the four subject areas that summed to 1 for each article. The figure shows a four way Venn Diagram~\cite{chen_venn_2011} illustrating the 
relative frequencies of publications from the Physical Sciences, Social Sciences, Life Sciences, and Health Sciences.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}
